---
title: "Homework 4"
format: html
editor: 
  markdown: 
    wrap: 72
---

# Homework 4

```{r}
library(httr)
library(jsonlite)
library(tidyverse)
```

## Task 1- Querying an API and a Tidy-Style Function

### Query API and Format Response to Return Article Info

```{r}

apiKey = "44c616340dfb495f90ed954af0159a95"



resp = httr::GET("https://newsapi.org/v2/everything?q=Iran&from=2026-05-13&language=en&pageSize=50&apiKey=44c616340dfb495f90ed954af0159a95")

#str(resp)
news = fromJSON(rawToChar(resp$content))

#str(news)

head(news$articles)

```

### Write Function to Query api

```{r}

getNews = function(querystr = NA, dateStr = Sys.Date(), apiKey = "44c616340dfb495f90ed954af0159a95"){
  if (!is.na(querystr)){
    query = paste(querystr,"&", sep = "")
  }
  else{
    query = ""
  }
  base_url = "https://newsapi.org/v2/everything?q="
  
  dateURL = paste0("from=",dateStr, sep = "")
  middleURL = "&language=en&pageSize=50&"
  apiURL = paste("apiKey=", apiKey, sep = "")
  
  url = paste(base_url, query,dateURL,middleURL, apiURL, sep = "")
  
  news = fromJSON(rawToChar(GET(url)$content))$articles
  
  return (news)
}

head(getNews())

head(getNews(querystr = "Aho" , dateStr = '2026-06-12'))

```

## Task 2

```{r}
coasters = read.csv("https://www4.stat.ncsu.edu/online/datasets/635USrollercoasters.csv")

head(coasters)
```

### Remove Boundary, Latitude, and Longitude variables

```{r}
coasters = coasters |> select(-c(Boundary, Latitude, Longitude))

str(coasters)
```

### Convert Data to Intended Types:

```{r}
coasters = coasters |> mutate(AgeGroup = as.factor(AgeGroup),
                              Scale = as.factor(Scale),
                              Type = as.factor(Type),
                              Design = as.factor(Design),
                              Year.Opened = as.factor(Year.Opened),
                              Census.Region = as.factor(Census.Region))


```

### Basic EDA

We will look how data is structured

```{r}
str(coasters)
```

### Assess Missiningness

```{r}
rate_na <- function(column){
  100*(sum(is.na(column)) / length(column))
}

na_counts = coasters |>
  summarize(across(everything(), rate_na))

na_counts

```

The following columns contain missing data:

-   TopSpeed (11.7% Missing)

-   MaxHeight (1.24% Missing)

-   Drop (10.2% Missing)

-   Length(5.2% Missing)

-   Duration (10.4% Missing)

All columns with missing data are numeric. We can do the following to
deal with the missing data: - Impute missing values (with means, group
means, etc) - Remove Observation with missing data

### Contingency Tables

#### 1-Way

```{r}

table(coasters$Type)
```

There are many more steel coasters than wood coasters in the dataset.

#### 2-Way

```{r}
table(coasters$Type, coasters$AgeGroup)

```

It looks like wood coasters have a more even distribution of age, while
steel coasters are mostly newer. This makes sense, as steel coasters
have not been around as long as wood coasters

#### 3-Way

```{r}
table(coasters$Type, coasters$AgeGroup, coasters$Census.Region)
```

The Northeast region has more old wood coasters than all of the other
regions by a large margin. Could this be because of all of the old
boardwalks in the northeast that had the first amusement parks?

### Conditional Two-Way Contingency Table:

Look at a 2-Way Contingency Table for Coaster Age and Region for Steel
coasters only.

```{r}
coasterTable = table(coasters$Type, coasters$AgeGroup, coasters$Census.Region)

(coasterTable[1, , ])
```

### 2-Way Contingency Table in deplyr

```{r}
coasters |>
  group_by(Type, AgeGroup) |>
  summarize(count = n()) |>
  pivot_wider(names_from = Type, values_from = count)
```

### Create Plots

#### Stacked Bar Graph

```{r}
g = ggplot(data = coasters, aes(x = Type, fill = AgeGroup))
g + geom_bar() + coord_flip()
```

#### Side By Side

```{r}
g = ggplot(data = coasters, aes(x = Type, fill = AgeGroup))
g + geom_bar(position = "dodge") 
```

### Numeric Data Summaries:

Note: for summary purposes, entries with blanks will be dropped

```{r}
coasterSummary = coasters |>
  drop_na() |>
  summarize(across(where(is.numeric), 
                   list("mean" = mean, "standardDev" = sd), 
                   .names = "{.fn}_{.col}"))

coasterSummary
```

### Summarize while filtering

Summary data for wood coasters only

```{r}
coasterSummaryWood = coasters |>
  drop_na() |>
  filter(Type == "Wood") |>
  summarize(across(where(is.numeric), 
                   list("mean" = mean, "standardDev" = sd), 
                   .names = "{.fn}_{.col}"))

coasterSummaryWood
```

### Numeric Summaries By Group (Region)

```{r}
coasterGroupSummary = coasters |>
  group_by(Census.Region) |>
  drop_na() |>
  summarize(across(where(is.numeric), 
                   list("mean" = ~ mean(.x, na.rm = TRUE), "standardDev" = ~ sd(.x, na.rm = TRUE)), 
                   .names = "{.fn}_{.col}"))

coasterGroupSummary
```

### Correlation between top speed and max height

```{r}
coasters |>
  drop_na() |>
  select(where(is.numeric)) |>
  cor()

```

### Distribution of Height by Type

#### Overlayed Histograms

```{r}
g <- ggplot(coasters |> drop_na(), aes(x = TopSpeed))
g + geom_histogram(alpha = 0.5, aes(fill = Type))
```

Wood coasters' top speed appears to be normally distributed. For steel
coasters, the distribution appears to be bimodal. This could be because
lots of new kids roller coasters could be steel.

#### Density Plot

```{r}
g <- ggplot(coasters |> drop_na(), aes(x = TopSpeed))
g + geom_density(alpha = 0.5, aes(fill = Type))
```

Wood and Steel coasters have a similar center, but the speed for steel
coasters has much more varation than that for wood coasters.

#### Box Plot

```{r}
g <- ggplot(coasters |> drop_na(), aes(x = TopSpeed))
g + geom_boxplot(alpha = 0.5, aes(fill = Type))
```

IQR is much wider for steel coasters than for wood coasters.

### Scatter Plots (Speed vs Height)

```{r}
g <- ggplot(coasters |> drop_na(), aes(x = TopSpeed, y = MaxHeight))
g + geom_point(aes(color = Type))
```

The relationship between speed and height seems to be consistent across
type. Steel coasters have higher heights and speed.

### Faceting For Generating Multiple Plots Across Region

```{r}
g <- ggplot(coasters |> drop_na(), aes(x = TopSpeed, y = MaxHeight))
g + geom_point(aes(color = Type)) + facet_grid(Census.Region ~ .)

```

Speed vs Height relationship holds across regions (this makes sense).
The West has the least wood coasters by far

### Speed vs Height Mutiple Facets (Region and Type)

```{r}
g <- ggplot(coasters |> drop_na(), aes(x = TopSpeed, y = MaxHeight))
g + geom_point() + facet_grid(Census.Region ~ Type)
```

When we look at the relationships between speed and height by the type
of coaster, it appears that the relationship between speed and height is
more exponential for steel coasters than wood coasters
