---
title: "Homework 4"
format: html
editor_options: 
  chunk_output_type: console
execute:
  warning: false
  message: false
---

#  Preamble
This homework practices with application programming interfaces (APIs) and exploratory data analyses (EDAs).

# Task 1: Querying an API and a Tidy-Style Function
This task makes use of the website newsapi.org. Having registered for a key using my NCSU email, we now do the following.

1. Use `GET()` from the `httr` package to return information about a topic I am interested in recently (i.e., within the past $30$ days, per the free account limit): the Ebola virus outbreak in the Democratic Republic of Congo (DRC) and Uganda.

```{r hw_4_task_1_response_1}
library(tidyverse)
library(dplyr)
library(httr)

ebola_news <- GET("https://newsapi.org/v2/everything?q=ebola&from=2026-05-15&language=en&pageSize=50&apiKey=e7e98c8f31914c04a5d15115a45ea6e3")

ebola_news
```

(For future reference, I cannot load multiple libraries in the same `library` call separated by commas.)

2. We now parse what was returned to recover information about article information.

```{r hw_4_task_1_response_2}
library(jsonlite)

ebola_parsed <- fromJSON(rawToChar(ebola_news$content))

str(ebola_parsed)
```

3. Lastly, we construct a function allowing a user to more easily query this API by `topic`, `earliest_date`, and `API_key`. For this problem, we will return a `vector` of the article titles.

```{r hw_4_task_1_response_3_part_1}
news_search <- function(topic,
                        earliest_date = "2026-05-16",
                        API_key = "e7e98c8f31914c04a5d15115a45ea6e3"){
  
  format_query <- paste0("https://newsapi.org/v2/everything?q=",
                         topic,
                         "&from=",
                         earliest_date,
                         "&language=en&pageSize=50&apiKey=",
                         API_key)
  
  news_query <- GET(format_query)
  
  news_parsed <- fromJSON(rawToChar(news_query$content))
  
  news_parsed$articles$title
}
```

First, let us query on "AI".

```{r hw_4_task_1_response_3_part_2}
news_search("AI")
```

Finally, let us query, within the past week, on the "NBA".

```{r hw_4_task_1_response_3_part_3}
news_search("NBA", "2026-06-08")
```

# Task 2: Summarizing Roller Coaster Data
For this task, we use the data located [here](https://www4.stat.ncsu.edu/online/datasets/635USrollercoasters.csv).

## Read In and Convert to Factors
First we read in the data, amending the column names to standard format.

```{r hw_4_task_2_response_01_part_1}
columns <- c("coaster", "park", "city", "state", "census_region",
             "year_opened", "age_group", "top_speed", "max_height", "scale",
             "type", "design", "drop", "length", "duration",
             "inversions", "number_of_inversions", "make", "latitude", "longitude",
             "boundary")

roller_coaster <- read_csv("https://www4.stat.ncsu.edu/online/datasets/635USrollercoasters.csv",
                  col_names = columns,
                  show_col_types = FALSE,
                  skip = 1,
  )
```

Then convert the appropriate variables (columns), viz., `census_region`, `age_group`, `scale`, `type`, and `design`, to factors; and convert an additional column, `inversions`, to `logi`.

```{r hw_4_task_2_response_01_part_2}
roller_coaster <- roller_coaster |>
  mutate(across(c(census_region, age_group, scale, type, design),
                as.factor))

for(i in 1:length(roller_coaster$inversions)){
  roller_coaster$inversions[i] <- ifelse(roller_coaster$inversions[i] == "Yes",
                                        "True",
                                        "False")
}

roller_coaster$inversions <- as.logical(roller_coaster$inversions)

str(roller_coaster)
```

Now we examine the data for missing values.

```{r hw_4_task_2_response_01_part_3}
summary(roller_coaster)
```

Of the $9$ numeric variables, $5$ have missing values: `top_speed`, `max_height`, `drop`, `length`, and `duration`. None of the `char` or `factor` variables appear to have missing values.

## Categorical Variables
Now we examine a few comparative tables and plots for the categorical data.

* First we create a one-way, two-way, and three-way contingency table in base `R` for the `factor` variables `census_region`, `scale`, and `age_group`, respectively.

```{r hw_4_task_2_response_02}
table(roller_coaster$census_region) |>
  addmargins()  # includes totals across rows, columns, and table levels

table(roller_coaster$census_region, roller_coaster$scale) |>
  addmargins()

three_way <- table(roller_coaster$census_region, roller_coaster$scale,
                   roller_coaster$age_group) |>
  addmargins()
```

  - In the first table, we observe the `South` has the most roller coasters according to this data set, at $223$.

  - In the second table, we observe that of the $160$ total `Thrill` roller coasters, the fewest occurs in the `West`, at $32$.

  - In the last table, we notice there aren't any (i.e., $0$) `Older` roller coasters in the `West` with a `Family` classification. Of the `Newest` roller coasters, the fewest occurs in the `Northeast` with `Kiddie` as `scale`.
    
* Next we create a conditional two-way table on the `Newest` roller coasters for `scale` and `census_region`.

```{r hw_4_task_2_response_03_part_1}
roller_coaster |>
  filter(age_group == "3:Newest") |>
  select(census_region, scale) |>
  table() |>
  addmargins()
```

Then we repeat this process by subsetting our previous three-way table.

```{r hw_4_task_2_response_03_part_2}
three_way[, , "3:Newest"]
```

* Then we create a two-way contingency table using `group_by` and `summarize` and `pivot_wider` to mimic the output of `table`.

```{r hw_4_task_2_response_04}
roller_coaster |>
  group_by(census_region, scale) |>
  summarize(count = n()) |>
  pivot_wider(names_from = scale, values_from = count)
```

* Now we begin to explore the data using graphs, starting with a stacked bar graph on `census_region` and `scale`.

```{r hw_4_task_2_response_05_part_1}
library(ggplot2)

ggplot(data = roller_coaster, aes(x = census_region, fill = scale)) +
  geom_bar() +
  labs(x = "Census Region", y = "Quantity",
       title = "Number of Roller Coasters",
       subtitle = "Stacked Bars") +
       theme(plot.title = element_text(hjust = 0.5),
             plot.subtitle = element_text(hjust = 0.5)) + # centers titles
       scale_fill_discrete("Scale") # Re-titles Legend
```

From this graph we can see `South` dominates the roller coaster scene.

Now we construct a side-by-side bar graph, this time using `census_region` and `age_group`

```{r hw_4_task_2_response_05_part_2}
ggplot(data = roller_coaster, aes(x = census_region, fill = age_group)) +
  geom_bar(position = "dodge") +
  labs(x = "Census Region", y = "Quantity",
       title = "Number of Roller Coasters",
       subtitle = "Side-by-Side Bar Plot") +
       theme(plot.title = element_text(hjust = 0.5),
             plot.subtitle = element_text(hjust = 0.5)) +
       scale_fill_discrete(name = "Age Group",
                           labels = c("Older", "Recent", "Newest")) # Re-titles Categories
```

For once, another category beats out `South`: `Northeast` has the highest quantity of `Older` roller coasters, which may make sense given the spread of industrialization across the US.

## Numeric Variables (and across groups)
For this section we will focus on the quantitative data from the roller coaster data set.

1. We begin by finding numeric summaries for `max_height` and `number_of_inversions`.

```{r hw_4_task_2_response_06_part_1}
roller_coaster |>
  summarize(across(c(max_height, number_of_inversions), 
                   list("mean" = ~ mean(.x, na.rm = TRUE),
                        "median" = ~ median(.x, na.rm = TRUE),
                        "std_dev" = ~ sd(.x, na.rm = TRUE),
                        "IQR" = ~ IQR(.x, na.rm = TRUE)),
                   .names = "{.fn}_{.col}")) |>
  simplify()
```

Evidently, a majority of roller coasters in the US have no inversions ("loop-de-loops"), as the median indicates. Since $\mu_{\text{max height}} > Q_{2_{\text{max height}}}$, we may conclude the `max_heights` is skewed right (positively skewed), i.e., in the direction of the mean.

Now we repeat the analyses, looking at only those roller coasters whose scale is `Extreme`.

```{r hw_4_task_2_response_06_part_2}
roller_coaster |>
  filter(scale == "Extreme") |>
  summarize(across(c(max_height, number_of_inversions), 
                   list("mean" = ~ mean(.x, na.rm = TRUE),
                        "median" = ~ median(.x, na.rm = TRUE),
                        "std_dev" = ~ sd(.x, na.rm = TRUE),
                        "IQR" = ~ IQR(.x, na.rm = TRUE)),
                   .names = "{.fn}_{.col}")) |>
  simplify()
```

Restricted to this `scale`, the mean and median `max_height`s are arppoximately $50$ feet (?) greater. The right-skewness of `max_height` is maintained, but within this group we note there are, on average, $2$ inversions.

2. We repeat the analyses, grouping by `census_region`.

```{r hw_4_task_2_response_07}
roller_coaster |>
  group_by(census_region) |>
  summarize(across(c(max_height, number_of_inversions), 
                   list("mean" = ~ mean(.x, na.rm = TRUE),
                        "median" = ~ median(.x, na.rm = TRUE),
                        "std_dev" = ~ sd(.x, na.rm = TRUE),
                        "IQR" = ~ IQR(.x, na.rm = TRUE)),
                   .names = "{.fn}_{.col}")) |>
  print(width = Inf) # prints all the columns
```

This shows the `mean_height` is consistently skewed right across the regions of the US, with the tallest roller coasters appearing to be in the `Midwest`, also with the highest variation. While the median `number_of_inversions` is still $0$, the `West` region appears to have the highest mean `number_of_inversions`.

3. We repeat the analyses, grouping by `census_region` and `scale`.

```{r hw_4_task_2_response_08}
roller_coaster |>
  group_by(census_region, scale) |>
  summarize(across(c(max_height, number_of_inversions), 
                   list("mean" = ~ mean(.x, na.rm = TRUE),
                        "median" = ~ median(.x, na.rm = TRUE),
                        "std_dev" = ~ sd(.x, na.rm = TRUE),
                        "IQR" = ~ IQR(.x, na.rm = TRUE)),
                   .names = "{.fn}_{.col}")) |>
  print(width = Inf)
```

This demonstrates that while the `Midwest` region appears to value `max_height` more than the other regions, the other regions appear to value `number_of_inversions` more than `Midwest`. Note there appears to be a correlation between the factor `Extreme` and `number_of_inversions`, as might be expected.

4. Next we craft a correlation matrix among all $9$ numeric variables.

```{r hw_4_task_2_response_09}
numeric_data <- roller_coaster[, sapply(roller_coaster, is.numeric)] # returns just the portion of the data that is numeric

cor(numeric_data, use = "complete.obs")
```

The strongest correlations appear between the `top_speed`, `max_height`, `drop`, and `length`, which from a physics standpoint makes sense, whereas the weakest correlations appear with the `latitude` and `longitude` variables. That makes sense because location should not have much impact on the other variables, and vice-versa.

5. 

6. 

7. 

8. 