---
title: "Homework 4"
format: html
editor_options: 
  chunk_output_type: console
execute:
  warning: false
  message: false
---

#  Preamble
This homework practices with application programming interfaces (APIs) and exploratory data analyses (EDAs).

# Task 1: Querying an API and a Tidy-Style Function
This task makes use of the website newsapi.org. Having registered for a key using my NCSU email, we now do the following.

1. Use `GET()` from the `httr` package to return information about a topic I am interested in recently (i.e., within the past $30$ days, per the free account limit): the Ebola virus outbreak in the Democratic Republic of Congo (DRC) and Uganda.

```{r hw_4_task_1_response_1}
library(tidyverse)
library(dplyr)
library(httr)

ebola_news <- GET("https://newsapi.org/v2/everything?q=ebola&from=2026-05-15&language=en&pageSize=50&apiKey=e7e98c8f31914c04a5d15115a45ea6e3")

ebola_news
```

(For future reference, I cannot load multiple libraries in the same `library` call separated by commas.)

2. We now parse what was returned to recover information about article information.

```{r hw_4_task_1_response_2}
library(jsonlite)

ebola_parsed <- fromJSON(rawToChar(ebola_news$content))

str(ebola_parsed)
```

3. Lastly, we construct a function allowing a user to more easily query this API by `topic`, `earliest_date`, and `API_key`. For this problem, we will return a `vector` of the article titles.

```{r hw_4_task_1_response_3_part_1}
news_search <- function(topic,
                        earliest_date = "2026-05-16",
                        API_key = "e7e98c8f31914c04a5d15115a45ea6e3"){
  
  format_query <- paste0("https://newsapi.org/v2/everything?q=",
                         topic,
                         "&from=",
                         earliest_date,
                         "&language=en&pageSize=50&apiKey=",
                         API_key)
  
  news_query <- GET(format_query)
  
  news_parsed <- fromJSON(rawToChar(news_query$content))
  
  news_parsed$articles$title
}
```

First, let us query on "AI".

```{r hw_4_task_1_response_3_part_2}
news_search("AI")
```

Finally, let us query, within the past week, on the "NBA".

```{r hw_4_task_1_response_3_part_3}
news_search("NBA", "2026-06-08")
```

# Task 2: Summarizing Roller Coaster Data
For this task, we use the data located [here](https://www4.stat.ncsu.edu/online/datasets/635USrollercoasters.csv).

## Read In and Convert to Factors
First we read in the data, amending the column names to standard format.

```{r hw_4_task_2_response_01_part_1}
columns <- c("coaster", "park", "city", "state", "census_region",
             "year_opened", "age_group", "top_speed", "max_height", "scale",
             "type", "design", "drop", "length", "duration",
             "inversions", "number_of_inversions", "make", "latitude", "longitude",
             "boundary")

roller_coaster <- read_csv("https://www4.stat.ncsu.edu/online/datasets/635USrollercoasters.csv",
                  col_names = columns,
                  show_col_types = FALSE,
                  skip = 1,
  )
```

Then convert the appropriate variables (columns), viz., `census_region`, `age_group`, `scale`, `type`, and `design`, to factors; and convert an additional column, `inversions`, to `logi`.

```{r hw_4_task_2_response_01_part_2}
roller_coaster <- roller_coaster |>
  mutate(across(c(census_region, age_group, scale, type, design),
                as.factor))

for(i in 1:length(roller_coaster$inversions)){
  roller_coaster$inversions[i] <- ifelse(roller_coaster$inversions[i] == "Yes",
                                        "True",
                                        "False")
}

roller_coaster$inversions <- as.logical(roller_coaster$inversions)

str(roller_coaster)
```

Now we examine the data for missing values.

```{r hw_4_task_2_response_01_part_3}
summary(roller_coaster)
```

Of the $9$ numeric variables, $5$ have missing values: `top_speed`, `max_height`, `drop`, `length`, and `duration`. None of the `char` or `factor` variables appear to have missing values.

## Categorical Variables
Now we examine a few comparative tables and plots for the categorical data.

* First we create a one-way, two-way, and three-way contingency table in base `R` for the `factor` variables `census_region`, `scale`, and `age_group`, respectively.

```{r hw_4_task_2_response_02}
table(roller_coaster$census_region) |>
  addmargins()  # includes totals across rows, columns, and table levels

table(roller_coaster$census_region, roller_coaster$scale) |>
  addmargins()

three_way <- table(roller_coaster$census_region, roller_coaster$scale,
                   roller_coaster$age_group) |>
  addmargins()
```

  - In the first table, we observe the `South` has the most roller coasters according to this data set, at $223$.

  - In the second table, we observe that of the $160$ total `Thrill` roller coasters, the fewest occurs in the `West`, at $32$.

  - In the last table, we notice there aren't any (i.e., $0$) `Older` roller coasters in the `West` with a `Family` classification. Of the `Newest` roller coasters, the fewest occurs in the `Northeast` with `Kiddie` as `scale`.
    
* Next we create a conditional two-way table on the `Newest` roller coasters for `scale` and `census_region`.

```{r hw_4_task_2_response_03_part_1}
roller_coaster |>
  filter(age_group == "3:Newest") |>
  select(census_region, scale) |>
  table() |>
  addmargins()
```

Then we repeat this process by subsetting our previous three-way table.

```{r hw_4_task_2_response_03_part_2}
three_way[, , "3:Newest"]
```

* Then we create a two-way contingency table using `group_by` and `summarize` and `pivot_wider` to mimic the output of `table`.

```{r hw_4_task_2_response_04}
roller_coaster |>
  group_by(census_region, scale) |>
  summarize(count = n()) |>
  pivot_wider(names_from = scale, values_from = count)
```

* Now we begin to explore the data using graphs, starting with a stacked bar graph on `census_region` and `scale`.

```{r hw_4_task_2_response_05_part_1}
library(ggplot2)

ggplot(data = roller_coaster, aes(x = census_region, fill = scale)) +
  geom_bar() +
  labs(x = "Census Region", y = "Quantity",
       title = "Number of Roller Coasters",
       subtitle = "Stacked Bars") +
       theme(plot.title = element_text(hjust = 0.5),
             plot.subtitle = element_text(hjust = 0.5)) + # centers titles
       scale_fill_discrete("Scale") # Re-titles Legend
```

From this graph we can see `South` dominates the roller coaster scene.

Now we construct a side-by-side bar graph, this time using `census_region` and `age_group`

```{r hw_4_task_2_response_05_part_2}
ggplot(data = roller_coaster, aes(x = census_region, fill = age_group)) +
  geom_bar(position = "dodge") +
  labs(x = "Census Region", y = "Quantity",
       title = "Number of Roller Coasters",
       subtitle = "Side-by-Side Bar Plot") +
       theme(plot.title = element_text(hjust = 0.5),
             plot.subtitle = element_text(hjust = 0.5)) +
       scale_fill_discrete(name = "Age Group",
                           labels = c("Older", "Recent", "Newest")) # Re-titles Categories
```

For once, another category beats out `South`: `Northeast` has the highest quantity of `Older` roller coasters, which may make sense given the spread of industrialization across the US.

## Numeric Variables (and across groups)
For this section we will focus on the quantitative data from the roller coaster data set.

1. We begin by finding numeric summaries for `max_height` and `number_of_inversions`.

```{r hw_4_task_2_response_06_part_1}
roller_coaster |>
  summarize(across(c(max_height, number_of_inversions), 
                   list("mean" = ~ mean(.x, na.rm = TRUE),
                        "median" = ~ median(.x, na.rm = TRUE),
                        "std_dev" = ~ sd(.x, na.rm = TRUE),
                        "IQR" = ~ IQR(.x, na.rm = TRUE)),
                   .names = "{.fn}_{.col}")) |>
  simplify()
```

Evidently, a majority of roller coasters in the US have no inversions ("loop-de-loops"), as the median indicates. Since $\mu_{\text{max height}} > Q_{2_{\text{max height}}}$, we may conclude the `max_heights` is skewed right (positively skewed), i.e., in the direction of the mean.

Now we repeat the analyses, looking at only those roller coasters whose scale is `Extreme`.

```{r hw_4_task_2_response_06_part_2}
roller_coaster |>
  filter(scale == "Extreme") |>
  summarize(across(c(max_height, number_of_inversions), 
                   list("mean" = ~ mean(.x, na.rm = TRUE),
                        "median" = ~ median(.x, na.rm = TRUE),
                        "std_dev" = ~ sd(.x, na.rm = TRUE),
                        "IQR" = ~ IQR(.x, na.rm = TRUE)),
                   .names = "{.fn}_{.col}")) |>
  simplify()
```

Restricted to this `scale`, the mean and median `max_height`s are arppoximately $50$ feet (?) greater. The right-skewness of `max_height` is maintained, but within this group we note there are, on average, $2$ inversions.

2. We repeat the analyses, grouping by `census_region`.

```{r hw_4_task_2_response_07}
roller_coaster |>
  group_by(census_region) |>
  summarize(across(c(max_height, number_of_inversions), 
                   list("mean" = ~ mean(.x, na.rm = TRUE),
                        "median" = ~ median(.x, na.rm = TRUE),
                        "std_dev" = ~ sd(.x, na.rm = TRUE),
                        "IQR" = ~ IQR(.x, na.rm = TRUE)),
                   .names = "{.fn}_{.col}")) |>
  print(width = Inf) # prints all the columns
```

This shows the `mean_height` is consistently skewed right across the regions of the US, with the tallest roller coasters appearing to be in the `Midwest`, also with the highest variation. While the median `number_of_inversions` is still $0$, the `West` region appears to have the highest mean `number_of_inversions`.

3. We repeat the analyses, grouping by `census_region` and `scale`.

```{r hw_4_task_2_response_08}
roller_coaster |>
  group_by(census_region, scale) |>
  summarize(across(c(max_height, number_of_inversions), 
                   list("mean" = ~ mean(.x, na.rm = TRUE),
                        "median" = ~ median(.x, na.rm = TRUE),
                        "std_dev" = ~ sd(.x, na.rm = TRUE),
                        "IQR" = ~ IQR(.x, na.rm = TRUE)),
                   .names = "{.fn}_{.col}")) |>
  print(width = Inf)
```

This demonstrates that while the `Midwest` region appears to value `max_height` more than the other regions, the other regions appear to value `number_of_inversions` more than `Midwest`. Note there appears to be a correlation between the factor `Extreme` and `number_of_inversions`, as might be expected.

4. Next we craft a correlation matrix among all $9$ numeric variables.

```{r hw_4_task_2_response_09}
numeric_data <- roller_coaster[, sapply(roller_coaster, is.numeric)] # returns just the portion of the data that is numeric

cor(numeric_data, use = "complete.obs")
```

The strongest correlations appear between the `top_speed`, `max_height`, `drop`, and `length`, which from a physics standpoint makes sense, whereas the weakest correlations appear with the `latitude` and `longitude` variables. That makes sense because location should not have much impact on the other variables, and vice-versa.

5. Next we create some graphs for the quantitative data.
  * We start with a histogram, here for `max_height` on the categorical variable `scale`.

```{r hw_4_task_2_response_10_part_1}
plot <- ggplot(roller_coaster |> drop_na(),
               aes(x = max_height, fill = scale)) +
  labs(x = "Maximum Height",
       y = "Count",
       title = "Height by Scale of Rollercoaster")

plot + geom_histogram() + labs(subtitle = "Histogram")
```

  This histogram demonstrates the `Extreme` scale tends to have greater `maximum_height` and `Kiddie` tends to have less, as expected.

 * Next we examine a (kernel) density plot for the same.
 
```{r hw_4_task_2_response_10_part_2}
plot + geom_density(alpha = 0.5) + labs(subtitle = "Density Plot")
```

  This reinforces the conclusion from the previous graph, although due to the opacity we can more easily see that the `Kiddie` tends less towards heights; `Family` is bimodal, likely for those who just want to enjoy the ride and those who want a more exciting experience; `Thrill` seems to be the go-between for `Kiddie` and `Extreme`, a sort of "average" roller coaster experience.

  * Then, we use a boxplot to compare the `top_speed` across `scale`.
  
```{r hw_4_task_2_response_10_part_3}
ggplot(roller_coaster |> drop_na(),
       aes(x = top_speed, fill = scale)) +
  labs(x = "Top Speed",
       title = "Top Speed by Scale of Rollercoaster",
       subtitle = "Boxplots") +
  geom_boxplot() +
  theme( # no y-axis information is needed
    axis.title.y = element_blank(), # Removes "mpg" text
    axis.text.y  = element_blank(), # Removes numbers/labels
    axis.ticks.y = element_blank()  # Removes tick lines
  )
```

  As expected, the `Kiddie` category seems to be the most consistent, with the most centered data and the least spread.  

6. Now we create two scatterplots.

  - The first will relate `top_speed` and `max_height` by `scale`.
  
```{r hw_4_task_2_response_11_part_1}
ggplot(roller_coaster |> drop_na(),
       aes(x = top_speed, y = max_height)) +
  labs(x = "Top Speed",
       y = "Maximum Height",
       title = "Top Speed v Height by Scale of Rollercoaster",
       subtitle = "Scatter Plot") +
  geom_point(aes(color = scale))
```

  We would think the `Kiddie` and `Family` rides would tend lower in `max_height` and `top_speed` with the `Thrill` and `Extreme` tending higher, and this scatter-plot affirms that. Moreover, there appears to be a quadratic relationship between speed and height.
  
  - Now we examine the height with the year the roller coaster was constructed, again across the `scale` variable.
  
```{r hw_4_task_2_response_11_part_2}
ggplot(roller_coaster |> drop_na(),
       aes(x = year_opened, y = max_height)) +
  labs(x = "Year Opened",
       y = "Maximum Height",
       title = "Year v Height by Scale of Rollercoaster",
       subtitle = "Scatter Plot") +
  geom_point(aes(color = scale))
```

  This graph confirms that as time went by and technology improved, so did the heights of the roller coasters, with `Kiddie` scale being thought of later and `Extreme` considered more as Boomers approached adulthood.

7. Next, we use faceting to examine how `design` impacted `max_height` and `top_speed`.

```{r hw_4_task_2_response_12}
ggplot(roller_coaster |> drop_na(),
       aes(x = top_speed, y = max_height)) +
  labs(x = "Top Speed",
       y = "Maximum Height",
       title = "Speed v Height by Design of Rollercoaster",
       subtitle = "Scatter Plot") +
  geom_point() +
  facet_grid(~ design)
```

The most common type of roller coaster is a `Sit Down` `design`, explaining why most of the data is in this `factor` level, and this also demonstrates that the faster and taller roller coaster tend to be more traditionally designed since the specialized roller coasters have lower speed and heights, comparatively.

8. Finally, we use faceting to examine how `scale` and `age_group` (of the roller coaster) impacts the `top_speed` and `number_of_iversions`.

```{r hw_4_task_2_response_13}
ggplot(roller_coaster |> drop_na(),
       aes(x = number_of_inversions, y = top_speed)) +
  labs(x = "Number of Inversions",
       y = "Top Speed",
       title = "Inversions v Speed by Age and Scale of Rollercoaster",
       subtitle = "Scatter Plot") +
  geom_point(position = position_jitter(width = 0.5)) +
  facet_grid(age_group ~ scale)
```

This shows the newer roller coasters exhibit the `Extreme` scale and tend towards higher numbers of inversions and that, perhaps counter-intuitively, the top speed has a modest, if small, effect on the number of inversions.