---
title: "Homework 4"
format: html
editor_options: 
  chunk_output_type: console
---

#  Preamble
This homework practices with application programming interfaces (APIs) and exploratory data analyses (EDAs).

# Task 1: Querying an API and a Tidy-Style Function
This task makes use of the website newsapi.org. Having registered for a key using my NCSU email, we now do the following.

1. Use `GET()` from the `httr` package to return information about a topic I am interested in recently (i.e., within the past $30$ days, per the free account limit): the Ebola virus outbreak in the Democratic Republic of Congo (DRC) and Uganda.

```{r hw_4_task_1_response_1}
library(tidyverse)
library(dplyr)
library(httr)

ebola_news <- GET("https://newsapi.org/v2/everything?q=ebola&from=2026-05-15&language=en&pageSize=50&apiKey=e7e98c8f31914c04a5d15115a45ea6e3")

ebola_news
```

(For future reference, I cannot load multiple libraries in the same `library` call separated by commas.)

2. We now parse what was returned to recover information about article information.

```{r hw_4_task_1_response_2}
library(jsonlite)

ebola_parsed <- fromJSON(rawToChar(ebola_news$content))

str(ebola_parsed)
```

3. Lastly, we construct a function allowing a user to more easily query this API by `topic`, `earliest_date`, and `API_key`. For this problem, we will return a `vector` of the article titles.

```{r hw_4_task_1_response_3_part_1}
news_search <- function(topic,
                        earliest_date = "2026-05-16",
                        API_key = "e7e98c8f31914c04a5d15115a45ea6e3"){
  
  format_query <- paste0("https://newsapi.org/v2/everything?q=",
                         topic,
                         "&from=",
                         earliest_date,
                         "&language=en&pageSize=50&apiKey=",
                         API_key)
  
  news_query <- GET(format_query)
  
  news_parsed <- fromJSON(rawToChar(news_query$content))
  
  news_parsed$articles$title
}
```

First, let us query on "AI".

```{r hw_4_task_1_response_3_part_2}
news_search("AI")
```

Finally, let us query, within the past week, on the "NBA".

```{r hw_4_task_1_response_3_part_3}
news_search("NBA", "2026-06-08")
```
